Manuscript progress report

### 2024-01-16 Docs/manuscript.Rmd
the second reference method (FOSR)

```{r}
load(here("Data/sim_data.RData"))
df_m <- sim_data[[1]]
df_m %>% filter(t>0.2 & t<=0.4)
```

We would like to use a second reference method for predictive performance comparision. In this madel, predictions on a specific interval are all made using the last few observations in the observed window. For example, if we are given observations on [0, 0.2] (j = 1...200), we may use L observations taken right before t=0.2($t_j: j=200, ... 200-(L-1)$) as time-fixed covariate to predict any future time.

@@

Under this framework, for each dataset we need to fit four models:

@@

## L = 1

@@

```{r fit_model, class.source='fold-show'}
# format training data
## given [0, 0.2], predictor would be
y_obs_max <- train_df %>% filter(window==1) %>% arrange(desc(t)) %>%
  group_by(id) %>%
  slice(L) %>% select(id, Y) %>% rename(yl = Y)
## to predict (0.2, 0.4), outcome:
df_pred_tr <- train_df %>% filter(window!=1) %>%
  left_join(y_obs_max, by = "id") %>% 
  mutate_at(vars(Y, yl), as.factor)

@@

>>>>

```{r}
load(here("Data/sim_data.RData"))
df_m <- sim_data[[1]]
```

We would like to use a second reference method for predictive performance comparision. In this model, predictions on a specific interval are all made using the last few observations in the observed window. For example, if we are given observations on [0, 0.2] (j = 1...200), we may use L observations taken right before t=0.2 ($t_j: j=200, ... 200-(L-1)$) as time-fixed covariate to predict any future time.

@@

Under this framework, for each dataset we need to fit four models (similar to the GLMMadaptive method):

@@

## Example of L = 1, given [0, 0.2] to predict (0.2, 1)

@@

```{r fit_model, class.source='fold-show'}
# format training data
## given [0, 0.2], predictor would be
y_obs_max <- train_df %>% filter(window==1) %>% arrange(desc(t)) %>%
  group_by(id) %>%
  slice(L) %>% select(id, Y) %>% rename(yl = Y)
## to predict (0.2, 1), outcome:
df_pred_tr <- train_df %>% filter(window!=1) %>%
  left_join(y_obs_max, by = "id") %>% 
  mutate_at(vars(Y, yl), as.factor)

# range(df_pred_tr$sind)

@@

## Example of prediction on one dataset (L = 1)

```{r}
# table(df_m$window)
train_df$window <- as.numeric(as.character(train_df$window))
test_df$window <- as.numeric(as.character(test_df$window))
test_df2 <- test_df

t1 <- Sys.time()
for(w in 1:4){
  # model fit
  # predictor (training set)
  y_obs_max <- train_df %>% filter(window==w) %>% arrange(desc(t)) %>%
    group_by(id) %>%
    slice(L) %>% select(id, Y) %>% rename(yl = Y)
  # outcome (training set)
  df_pred_tr <- train_df %>% filter(window > w) %>%
    left_join(y_obs_max, by = "id") %>% 
    mutate_at(vars(Y, yl), as.factor)
  # model 
  fit_gen_fosr <- bam(Y ~ s(t, bs="cc", k=20) + 
                        s(t, bs="cc", k=20, by = yl),
                   family = binomial, data=df_pred_tr, 
                   method = "fREML",
                   discrete = TRUE)

  # prediction
  # predictor (testing set)
  y_obs_max_te <- test_df %>% 
                filter(window==w) %>% arrange(desc(t)) %>% group_by(id) %>%
                slice(L) %>% select(id, Y) %>% rename(yl = Y)
  # outcome (testing set)
  df_pred_te <- test_df %>% filter(window > w) %>%
    left_join(y_obs_max_te, by = "id") %>% 
    mutate_at(vars(Y, yl), as.factor)

  # predict using the FOSR model fit above
  pred_name <- paste0("pred_w", w)
  test_df[, pred_name] <- NA
  test_df[test_df$window>w, pred_name] <- predict(fit_gen_fosr, newdata = df_pred_te, type = "link")
}
t2 <- Sys.time()
# t2-t1

```{r}
test_df %>%
  mutate_at(vars(eta_i, pred_w1, pred_w2, pred_w3, pred_w4), function(x){exp(x)/(1+exp(x))}) %>%
  mutate_at(vars(Y), function(x){as.numeric(as.character(x))}) %>%
  filter(id %in% rand_id) %>%
  ggplot()+
  geom_point(aes(x=t, y=Y), size = 0.5)+
  geom_line(aes(x=t, y=pred_w1, col = "Up to 0.1"), na.rm=T)+
  geom_line(aes(x=t, y=pred_w2, col = "Up to 0.2"), na.rm=T)+
  geom_line(aes(x=t, y=pred_w3, col = "Up to 0.3"), na.rm=T)+
  geom_line(aes(x=t, y=pred_w4, col = "Up to 0.4"), na.rm=T)+
  geom_line(aes(x=t, y=eta_i, col = "ture"))+
  facet_wrap(~id)+
  labs(title = "Out-of-sample prediction")
```

## Example of prediction on one dataset (L = 5)

```{r}
t1 <- Sys.time()
for(w in 1:4){
  # model fit
  # predictor (training set)
  y_obs_max <- train_df %>% filter(window==w) %>% group_by(id) %>%
    slice_max(t, n=5) %>% select(id, Y, sind) %>%
    mutate(name = sind-min(sind)+1) %>%
    pivot_wider(id_cols = id, values_from = Y, names_from = "name", names_prefix = "yl")
  # outcome (training set)
  df_pred_tr <- train_df %>% filter(window > w) %>%
    left_join(y_obs_max, by = "id") %>% 
    mutate_at(vars(Y, starts_with("yl")), as.factor) 
  fit_gen_fosr <- bam(Y ~ s(t, bs="cc", k=20) + 
                        s(t, bs="cc", k=20, by = yl5)+
                        s(t, bs="cc", k=20, by = yl4)+
                        s(t, bs="cc", k=20, by = yl3)+
                        s(t, bs="cc", k=20, by = yl2)+
                        s(t, bs="cc", k=20, by = yl1),
                   family = binomial, data=df_pred_tr, 
                   method = "fREML",
                   discrete = TRUE)

  # prediction
  # predictor (testing set)
  y_obs_max_te <- test_df2 %>% filter(window==w) %>% group_by(id) %>%
    slice_max(t, n=5) %>% select(id, Y, sind) %>% 
    mutate(name = sind-min(sind)+1) %>%
    pivot_wider(id_cols = id, values_from = Y, names_from = name, names_prefix = "yl")
  # outcome (testing set)
  df_pred_te <- test_df2 %>% filter(window > w) %>%
    left_join(y_obs_max_te, by = "id") %>% 
    mutate_at(vars(Y, starts_with("yl")), as.factor)

  # predict using the FOSR model fit above
  pred_name <- paste0("pred_w", w)
  test_df2[, pred_name] <- NA
  test_df2[test_df2$window>w, pred_name] <- predict(fit_gen_fosr, newdata = df_pred_te, type = "link")
}
t2 <- Sys.time()
# t2-t1

```{r}
test_df2 %>%
  mutate_at(vars(eta_i, pred_w1, pred_w2, pred_w3, pred_w4), function(x){exp(x)/(1+exp(x))}) %>%
  mutate_at(vars(Y), function(x){as.numeric(as.character(x))}) %>%
  filter(id %in% rand_id) %>%
  ggplot()+
  geom_point(aes(x=t, y=Y), size = 0.5)+
  geom_line(aes(x=t, y=pred_w1, col = "Up to 0.1"), na.rm=T)+
  geom_line(aes(x=t, y=pred_w2, col = "Up to 0.2"), na.rm=T)+
  geom_line(aes(x=t, y=pred_w3, col = "Up to 0.3"), na.rm=T)+
  geom_line(aes(x=t, y=pred_w4, col = "Up to 0.4"), na.rm=T)+
  geom_line(aes(x=t, y=eta_i, col = "ture"))+
  facet_wrap(~id)+
  labs(title = "Out-of-sample prediction")
```

This method is very, very fast, takes 10 seconds or so, much faster than both fGFPCA and GLMMadaptive. The performance lies in between them for most cases, sometimes worse than GLMMadaptive.

Do we really want to use the lag-1 model?

```{r ise_fosr}
tb_ies_l5 <- test_df2 %>%
    mutate(err1 = (pred_w1-eta_i)^2,
           err2 = (pred_w2-eta_i)^2,
           err3 = (pred_w3-eta_i)^2,
           err4 = (pred_w4-eta_i)^2) %>%
    select(id, t, window, starts_with("err")) %>% 
    group_by(window, id) %>% 
    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
    group_by(window) %>% 
    summarize_at(vars(err1, err2, err3, err4), mean) %>%
    filter(window != 1) %>%
  select(starts_with("err"))

tb_ies_l1 <- test_df %>%
    mutate(err1 = (pred_w1-eta_i)^2,
           err2 = (pred_w2-eta_i)^2,
           err3 = (pred_w3-eta_i)^2,
           err4 = (pred_w4-eta_i)^2) %>%
    select(id, t, window, starts_with("err")) %>% 
    group_by(window, id) %>% 
    summarise_at(vars(err1, err2, err3, err4), sum) %>% 
    group_by(window) %>% 
    summarize_at(vars(err1, err2, err3, err4), mean) %>%
    filter(window != 1) %>%
  select(starts_with("err"))

colnames(tb_ies_l1) <- colnames(tb_ies_l5) <- c("0.2", "0.4", "0.6", "0.8")
```

```{r}
data.frame(Window = c("(0.2,0.4]", "(0.4,0.6]", "(0.6,0.8]", "(0.8, 1.0]"),
           tb_ies_l1, tb_ies_l5, check.names = F) %>%
  kable(digits = 3, caption = "Integrated squared error", booktabs = T) %>%
  kable_styling(full_width = F) %>% 
  add_header_above(c(" "=1, "L = 1" = 4, "L = 5" = 4)) %>%
  add_header_above(c(" "= 1, "Observation track" = 8))
```
